---
title: "Pedigree uncertainty in LCA: a worked example"
author: "lcaStats · Harper Food Innovation: Digital"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pedigree uncertainty in LCA: a worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  dpi = 150
)
```

## Introduction

Uncertainty is an unavoidable feature of life cycle assessment (LCA). The 
pedigree matrix, introduced by @weidema1996 and formalised for ecoinvent v3 
by @weidema2013, provides a systematic method for converting expert judgements 
about data quality into quantitative uncertainty estimates. Each inventory 
exchange is rated across five **Data Quality Indicators (DQIs)**:

| Indicator | Question |
|---|---|
| **Reliability** | How well does the source measure what we need? |
| **Completeness** | How fully does the data cover the relevant population? |
| **Temporal correlation** | How recent is the data? |
| **Geographical correlation** | How geographically representative is the data? |
| **Technology correlation** | How technologically representative is the data? |

Scores range from **1** (best) to **5** (worst). Each score maps to an 
additional variance term in log-space ($\sigma^2_{ln}$), which is summed with 
the exchange's basic uncertainty to give a combined lognormal distribution.

This vignette walks through a realistic example using N₂O field emissions 
from an agricultural system — a common hotspot in food system LCAs.

---

## Setup

```{r load}
# Source the pedigree module
# If the package is installed: library(lcaStats)
source("../R/pedigree.R")
library(ggplot2)
library(dplyr)
```

---

## 1. A single exchange

Consider a direct N₂O field emission factor from a UK arable study, scored 
as follows:

- **Reliability (3):** Data from a proxy measurement (IPCC Tier 1 approach), 
  not a direct site measurement.
- **Completeness (2):** Based on multiple UK sites covering the main 
  production regions.
- **Temporal correlation (3):** Data are 8–12 years old relative to the 
  study reference year.
- **Geographical correlation (1):** UK-specific data, perfectly matched.
- **Technology correlation (2):** Conventional arable, closely matching the 
  system under study.

```{r single}
result <- scores_to_gsd2(
  reliability              = 3,
  completeness             = 2,
  temporal_correlation     = 3,
  geographical_correlation = 1,
  technology_correlation   = 2,
  basic_var                = 0.0006
)

pedigree_summary(result, exchange_name = "N2O field emission — UK arable")
```

The **GSD²** is the ecoinvent standard for expressing lognormal uncertainty. 
A GSD² of `r round(result$gsd2, 3)` indicates relatively modest uncertainty 
— the 97.5th percentile is approximately `r round(result$gsd, 2)` times the 
median.

---

## 2. Monte Carlo sampling

Using the lognormal parameters, we can draw a sample distribution for this 
exchange. Suppose the inventory value (arithmetic mean) is 
**0.014 kg N₂O-N per kg N applied**.

```{r mc}
params  <- lognormal_params(mean = 0.014, sigma_ln = result$sigma_ln)
set.seed(123)
samples <- rlnorm(50000, meanlog = params["mu_ln"], sdlog = params["sigma_ln"])

cat(sprintf(
  "Mean:         %.5f\nMedian:       %.5f\n2.5th pct:    %.5f\n97.5th pct:   %.5f\n",
  mean(samples), median(samples),
  quantile(samples, 0.025), quantile(samples, 0.975)
))
```

```{r mc-plot, fig.cap = "Lognormal distribution for the N2O field emission factor (n = 50,000 draws). Dashed line = median; dotted lines = 2.5th and 97.5th percentiles."}
df_samples <- data.frame(value = samples)
p025 <- quantile(samples, 0.025)
p975 <- quantile(samples, 0.975)
p50  <- median(samples)

ggplot(df_samples, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 80, fill = "#2e7d52", alpha = 0.75, colour = NA) +
  geom_vline(xintercept = p50,  linetype = "dashed", colour = "#111",
             linewidth = 0.9, show.legend = TRUE) +
  geom_vline(xintercept = p025, linetype = "dotted", colour = "#b55",
             linewidth = 0.9) +
  geom_vline(xintercept = p975, linetype = "dotted", colour = "#b55",
             linewidth = 0.9) +
  labs(
    x = expression("N"[2]*"O-N emission factor (kg kg N"^{-1}*")"),
    y = "Density",
    title = "Lognormal uncertainty distribution",
    subtitle = "N\u2082O field emission \u2014 UK arable system"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

---

## 3. Variance breakdown — what drives the uncertainty?

A key output of the pedigree approach is understanding which indicators 
contribute most to total uncertainty. This informs targeted data collection 
efforts.

```{r breakdown-plot, fig.cap = "Contribution of each uncertainty source to total variance (sigma^2_ln). Improving the highest-contributing indicator gives the greatest reduction in uncertainty."}
sources   <- c("Basic uncertainty", names(INDICATOR_LABELS))
labels    <- c("Basic uncertainty", as.character(INDICATOR_LABELS))
variances <- c(result$basic_var, result$indicator_vars)
pct       <- variances / result$combined_var * 100

df_break <- data.frame(
  source    = factor(labels, levels = rev(labels)),
  variance  = variances,
  pct       = pct
)

ggplot(df_break, aes(x = pct, y = source)) +
  geom_col(fill = "#1565c0", alpha = 0.8, width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", pct)),
            hjust = -0.1, size = 3.5, colour = "#333") +
  scale_x_continuous(
    limits = c(0, max(pct) * 1.2),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    x = "% of total variance",
    y = NULL,
    title = "Pedigree variance breakdown",
    subtitle = "N\u2082O field emission \u2014 UK arable system"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())
```

---

## 4. Batch processing — a multi-exchange inventory

In practice, we apply the pedigree approach across all exchanges in an 
inventory. Here we demonstrate with a small example inventory covering 
key inputs and outputs of a wheat production system.

```{r batch}
inventory <- data.frame(
  exchange = c(
    "N2O field emissions",
    "Diesel — field operations",
    "Nitrogen fertiliser (ammonium nitrate)",
    "Pesticide application",
    "Wheat grain yield"
  ),
  reliability              = c(3L, 2L, 1L, 3L, 2L),
  completeness             = c(2L, 2L, 2L, 3L, 1L),
  temporal_correlation     = c(3L, 2L, 1L, 3L, 2L),
  geographical_correlation = c(1L, 1L, 1L, 2L, 1L),
  technology_correlation   = c(2L, 2L, 1L, 3L, 2L),
  basic_var                = rep(0.0006, 5)
)

results <- batch_from_dataframe(inventory)

# Show key uncertainty columns
results |>
  select(exchange, combined_var, gsd2, sigma_ln) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

```{r batch-plot, fig.cap = "GSD^2 across inventory exchanges. Higher values indicate greater uncertainty. Exchanges above the dashed line (GSD^2 = 1.005) would be priority targets for improved data collection."}
ggplot(results, aes(x = gsd2, y = reorder(exchange, gsd2))) +
  geom_col(fill = "#2e7d52", alpha = 0.75, width = 0.6) +
  geom_vline(xintercept = 1.005, linetype = "dashed",
             colour = "#b55", linewidth = 0.8) +
  geom_text(aes(label = round(gsd2, 4)),
            hjust = -0.1, size = 3.5, colour = "#333") +
  scale_x_continuous(limits = c(1, max(results$gsd2) * 1.15)) +
  labs(
    x = "GSD²",
    y = NULL,
    title = "Pedigree uncertainty by exchange",
    subtitle = "Wheat production system — illustrative inventory"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())
```

---

## 5. Sensitivity: what if we improve temporal correlation?

A common practical question is: *what would be gained by collecting more 
recent data?* We can answer this by varying a single indicator score and 
observing the change in GSD².

```{r sensitivity}
scores_range <- 1:5

sensitivity <- lapply(scores_range, function(tc_score) {
  r <- scores_to_gsd2(
    reliability              = 3,
    completeness             = 2,
    temporal_correlation     = tc_score,
    geographical_correlation = 1,
    technology_correlation   = 2,
    basic_var                = 0.0006
  )
  data.frame(temporal_score = tc_score, gsd2 = r$gsd2, sigma_ln = r$sigma_ln)
}) |> bind_rows()

sensitivity
```

```{r sensitivity-plot, fig.cap = "Effect of temporal correlation score on GSD^2. Each one-point improvement in the score reduces the combined uncertainty."}
ggplot(sensitivity, aes(x = temporal_score, y = gsd2)) +
  geom_line(colour = "#1565c0", linewidth = 1) +
  geom_point(colour = "#1565c0", size = 3) +
  geom_point(data = sensitivity[sensitivity$temporal_score == 3, ],
             colour = "#b55", size = 4, shape = 18) +
  annotate("text", x = 3.15, y = sensitivity$gsd2[3] + 0.0003,
           label = "Current score (3)", hjust = 0, size = 3.5, colour = "#b55") +
  scale_x_continuous(breaks = 1:5) +
  labs(
    x     = "Temporal correlation score",
    y     = "GSD²",
    title = "Sensitivity to temporal correlation",
    subtitle = "N\u2082O field emission \u2014 all other scores held constant"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Moving from score 3 to score 1 (using data from within the reference year) 
would reduce GSD² from `r round(sensitivity$gsd2[3], 4)` to 
`r round(sensitivity$gsd2[1], 4)` — a worthwhile improvement, though in 
this case the temporal indicator is not the dominant source of uncertainty 
(see Section 3).

---

## Summary

This vignette demonstrated:

1. Converting DQI scores for a single exchange into lognormal uncertainty 
   parameters using `scores_to_gsd2()`.
2. Drawing Monte Carlo samples with `rlnorm()` for propagation.
3. Visualising the variance breakdown to identify priority indicators.
4. Applying the approach in batch across a multi-exchange inventory.
5. Sensitivity analysis to quantify the benefit of data improvement.

The interactive Shiny app (`inst/shiny/app.R`) provides a point-and-click 
interface for the same workflow.

---

## References

```{=html}
<div id="refs"></div>
```

::: {#refs}
Frischknecht, R., Jungbluth, N., Althaus, H.-J., Doka, G., Dones, R., Heck,
T., Hellweg, S., Hischier, R., Nemecek, T., Rebitzer, G., & Spielmann, M.
(2005). The ecoinvent database: Overview and methodological framework.
*International Journal of Life Cycle Assessment*, 10(1), 3–9.
https://doi.org/10.1065/lca2004.10.181.1

Weidema, B. P., & Wesnæs, M. S. (1996). Data quality management for life 
cycle inventories — an example of using data quality indicators. 
*Journal of Cleaner Production*, 4(3–4), 167–174.
https://doi.org/10.1016/S0959-6526(96)00043-1

Weidema, B. P., Bauer, C., Hischier, R., Mutel, C., Nemecek, T., Reinhard,
J., Vadenbo, C. O., & Wernet, G. (2013). *Overview and methodology: Data
quality guideline for the ecoinvent database version 3.* Ecoinvent Report
1(v3). The ecoinvent Centre, St. Gallen.
:::
